---
title: "Reproducible Research: Peer Assessment 1"
output: html_document
author: "Amir Pournasserian"
date: "Sunday, September 07, 2014"
keep_md: true
---
## Loading and preprocessing the data
First of all, I should read raw dataset from the source file. As we know, the type of *date* column will be charachter, so I change the type to *Date*. By default, the other column's type are *int*.
```{r}
raw_data <- read.csv("activity.csv")
raw_data$date <- as.Date(as.character(raw_data$date))
```

## What is mean total number of steps taken per day?
```{r}
# tspd includes total (sum) number of steps taken per day 
tspd <- aggregate(steps ~ date, raw_data, sum)
hist(tspd$steps, xlab = "Steps", 
     main = "Total Number of Steps per Day")
```

####Mean  
The ***mean*** total number of steps taken per day is ***`r format(round(mean(tspd$steps), 0))`***.  

####Median
The ***median*** total number of steps taken per day is ***`r format(round(median(tspd$steps), 0))`***.

## What is the average daily activity pattern?
```{r}
# mspi includes the mean steps per interval
mspi <- aggregate(steps ~ interval, raw_data, mean)
plot(mspi, type = "l", xlab = "Intervals", 
     ylab = "Steps (Average)", 
     main = "Average Number of Steps (across all days)")
```

The 5-minute interval which the ***maximum*** number of steps on average across all the days is 
***`r mspi[mspi$steps == max(mspi$steps),1]`***.

## Imputing missing values

####Missing values
The total number of ***missing values*** in dataset is 
***`r length(raw_data[is.na(raw_data$steps),1])`***.

####Devising for missing value replacement
I replace the missing values with the mean for that 5-minute interval. 
Here are the steps which I applied for this purpose:

1. Join the raw dataset with the mean dataset. 
2. Find the index of rows which contain missing values.
3. Replace the missing column with mean data in the same index.

```{r}
devised_data <- merge(x = raw_data, y = mspi, by = "interval")
colnames(devised_data) <- c("interval", "steps", "date", "stepsMean")
ind <- which(is.na(devised_data$steps), arr.ind = TRUE)
devised_data[ind, 2] <- devised_data[ind, 4]
```

The new dataset that is equal to the original dataset but with the missing data filled in is called *devised_data*.

```{r}
dev_tspd <- aggregate(steps ~ date, devised_data, sum)
hist(dev_tspd$steps, xlab = "Steps", 
     main = "Total Number of Steps per Day")
```
  
####Devised Mean  
The ***mean*** total number of steps taken per day is ***`r format(round(mean(dev_tspd$steps), 0))`***.  

####Devised Median
The ***median*** total number of steps taken per day is ***`r format(round(median(dev_tspd$steps), 0))`***.


## Are there differences in activity patterns between weekdays and weekends?
```{r}
devised_data$is_weekend <- (weekdays(devised_data$date) %in% c("Saturday", "Sunday"))
devised_data$datetype <- "weekday"
devised_data$datetype[devised_data$is_weekend == TRUE] <- "weekend"
devised_data$datetype <- as.factor(devised_data$datetype)
view_data <- aggregate(steps ~ interval, devised_data, mean)
```

```{r}
library(lattice)
steps <- devised_data$steps
interval <- devised_data$interval
datetype <- devised_data$datetype
xyplot(steps ~ interval|datetype, 
       data = view_data, 
       type = "l",
       main="", 
       ylab = "Number of steps",
       xlab="Interval",
       layout = c(1,2))
```

